\Wxxsir methods to estimate $\rr$ in the SIR model

### 2018-07-03 paper/Version12/draft_v12.tex
finally on git

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
%\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}% http://ctan.org/pkg/graphicx
\usepackage{systeme}
%% tables
\usepackage{booktabs}
\usepackage[table,xcdraw]{xcolor}
\usepackage{float}

@@

\section{Introduction}\label{sec:intro}
What has been called ``arguably the most important quantity in the study of epidemics'' \citep{Heesterbeek2002},  $\mathcal{R}_0$, the reproduction number (by convention pronounced ``R-naught'') is a persistently elusive quantity for epidemic modelers to estimate.  As defined by \citet{anderson1992}, $\rr$ is the ``the average number of secondary infections produced when one infected individual is introduced into a host population where everyone is susceptible.''  $\rr$, in some ways, summarizes an entire outbreak of a disease; it  is used to assess whether a disease epidemic will occur.  Additionally, it describes what percentage of the population needs to be vaccinated to avoid such an epidemic, roughly $1-\rr^{-1}$ \citep{anderson1992}).  Despite a clear definition of $\rr$, epidemiologists have struggled to create a standard  estimator for $\rr$  \citep{hethcote2000}.  A major issue in estimating $\rr$ is that the quantity is a \textit{property of the model}, meaning that $\rr$ is dependent not only on the usual noise that comes with statistical modeling but also on a variety of assumptions on how researchers assume a disease is transmitted through a population.

To elaborate on $\rr$ being the property of the model, we first need to introduce the concept of the ``SI-framework,'' which was pioneered by Kermack and McKendrick in the 1920s \citep{getz2006}.   Here, S and I are known as compartments where S stands for ``susceptible'' and I for ``infectious.'' Infectious disease models then specify how individuals move from the S compartment to the I compartment.  Often, ancillary compartments are added such as R, which stands for recovered or dead, or E, which stands for exposed but not yet infectious.  The estimate of $\rr$ from a SIR model will not be directly comparable to the estimate of $\rr$ from a SEIR model, for example.  The difficulties of estimating $\rr$ are summarized and expanded upon by \cite{li2011}.   As a result of $\rr$ being a property of the model, we limit our review to methods for estimating $\rr$ for the SIR model introduced by \cite{Kermack700}.

@@

We review \wxxsir methods for estimating $\rr$.  We note that these methods are not exhaustive but are chosen due to their impact and use in epidemiology.  We find that these methods also help to highlight the problems discussed above.  In addition to making a point estimate of $\rr$, we are particularly interested in the hypothesis  $H_0: \rr > 1$, as a value greater than 1 indicates an outbreak of a disease.  As such, we discuss techniques for estimating CIs for each of the different methods.  These CI estimates are the delta method, block bootstrap, and the posterior distribution.

In the following sections, we present a series of simulations in which we examine each of our $\rr$ estimation methods.  Within these simulations, we vary the number of observations used to estimate $\rr$, infection and recovery rates, the total population size, initial SI ratio, and several assumptions about the noise.

We find that...

Finally, we discuss our recommendations in estimating $\rr$ for the SIR model, and we comment on how these recommendations extend to more complex models and how many of the problems with estimating $\rr$  become exacerbated with the increase in complexity.

The rest of this manuscript is organized as follows.  In Section \ref{sec:r0}, we briefly discuss the origin of $\rr$.  In Section \ref{sec:sir-intro}, we introduce the SIR model as described by \cite{Kermack700}.  In Section \ref{sec:methods}, we overview the \wxxsir methods for estimating $\rr$. In Section \ref{sec:ci}, we discuss how to estimate CIs for each of the \wxxsir methods.  Following in Section \ref{sec:sim-res}, we present the results of our simulations.  Finally, in Section \ref{sec:discussion}, we provide final analysis of estimating $\rr$ for the SIR model and offer comments on how these conclusions extend to more complex models.

@@

Other difficulties in estimating $\rr$ arise from model assumptions on how randomness enters the model and sensitivity of the estimate.  For example, the SIR model is unique in that two of its compartments, S and R are monotonic.  The number of susceptibles is non-increasing and the number of susceptibles is non-decreasing.  Models should enforce this monotonicity of the two compartments.  As a consequence, this restricts assumptions on how randomness enters the model.  For example, Gaussian error, a common assumption in statistics and machine learning, becomes highly suspect due to the non-symmetrical nature of the noise imposed by the monotonic $S$ and $R$ compartments.  Without assuming Gaussian noise, it becomes much more difficult to guarantee properties such as unbiased and consist estimates of $\rr$.

@@

% A final difficulty in estimating $\rr$ arises from the data.  Ideally, we would have counts of the number of susceptibles, infectious, and recovered at each time step.  However, sometimes only the incidence of the disease is available, where we actually need the prevalence.  Then, the number of susceptibles is back-estimated from the number of infectious which requires an estimate of the number of susceptibles, and even the US Census is known to be at least 10\% off in both directions of undercounting and overcounting (CITE).  Finally, the distinction between susceptible and infectious and infectious and recovered may not be clear.  As asymptomatic individuals may infect others and be mistaken to be in either the susceptible or infectious compartment.

As a result of these difficulties, accurately estimating $\rr$ is a monumentous task.  We overview \wxxsir methods introduced by a number of researchers used to solve some of the problems described above.

@@

In this particular  SIR model, $N$, the number of individuals is constant, as we see no arrows pointing away from the three nodes in Figure \ref{fig::sir}. Simple adaptations of the SIR model can include birth and death rates, which may correspondingly change the derivation of $\rr$ (for further discussion, see Section \ref{sec:discussion}). The parameters  $\beta$ and $\gamma$ have epidemiological meaning.  Here, $\beta$ is the average infection rate and $\gamma$ is the average recovery rate.  The movement of individuals from one compartment to another is represented through the ordinary differential equations below.  Both $\beta$ and $\gamma$ are constrained to be between $0$ and $1$.  For the remainder of this paper, we use $X$, $Y$, and $Z$ for stand-ins for the ODEs for $S$, $I$, and $R$, respectively, to avoid any confusion with $\rr$.
\begin{align}
\systeme{\frac{dX}{dt} = -\frac{\beta XY}{N}, \frac{dY}{dt} = \frac{\beta XY}{N} - \gamma Y, \frac{dZ}{dt} = \gamma Y}. \label{eq:sir}
\end{align}
In words, susceptible individuals become infected at a rate that is proportional to the percentage of infected individuals multiplied by $\beta$, the infection rate, and the number of susceptible individuals.  Infectious individuals recover at a rate of $\gamma$ multiplied by the number of infected individuals.

@@

In summary, when we discuss estimators for $\rr$ for the SIR model, we mean to say we are forming an estimator of $\rr$ from the given set of data in Eq. \ref{eq:sir-noise},
\begin{align*}
  \textnormal{Data} &= \left \{\left (X_{obs}(t), Y_{obs}(t), Z_{obs}(t) \right ) : t=0, 1, \dots, T\right \}, \\
  \hat{\rr} &= m(Data),
\end{align*}
where $m$ is a function of the data.

Good properties of an estimator may include being (1) being a consistent estimator
\begin{align*}
  \hat{\rr} \overset{P}{\to} 0 \textnormal{ as } T\to \infty
\end{align*}
or (2) converging to a known probability distribution,
\begin{align*}
\hat{\rr} \leadsto F.  
\end{align*}
As we are interested in the hypothesis
\begin{align*}
  H_0:\;& \rr > 1 \\
  H_A:\;& \rr \le 1
\end{align*}
Then for some $\alpha \in [0,1]$, we would like to know for some $a, b >0$
\begin{align*}
P(a \le \hat{\rr} \le b) = 1 - \alpha,  
\end{align*}
the second property is particularly desirable.  At any rate, we would like to be able to estimate $a$ and $b$ for a given $\alpha$-level.

@@

\subsection{Least Squares ($\beta$, $\gamma$) (LS)}\label{least-squares-beta-gamma}
The first approach to estimate $\rr$ in the SIR model is to minimize the joint sum of squares as seen in approaches similar to \cite{malik2008} and \cite{canto2017}.  In this method, the least squares estimate of $\rr$ is formed from the data collected at each time point to estimate $\hat{\beta}$ and $\hat{\gamma}$ and uses the  subsequent plug-in estimator found in Equation \ref{eq:sirls} to estimate $\rr$.  In particular, we find

\begin{align*}
(\hat{\beta}, \hat{\gamma} )&=\arg \min_{\beta, \gamma} \sum_{t} \left [ \left (X_{obs}(t) - X(t)\right )^2 + \left ( Z_{obs}(t) - Z(t) \right )^2 \right ]
\end{align*}
Then  the estimate for $\rr$ is given by Equation \ref{eq:sirls},
\begin{align}\label{eq:sirls}
  \hat{\rr}= \frac{\hat{\beta}}{\hat{\gamma}}.
\end{align}

This method is sometimes used to estimate $\hat{\beta}$ and $\hat{\gamma}$ by using a grid search but can be done with more sophisticated minimization algorithms.  Due to the ODE format of the equations, we can only use minimization algorithms which do not rely on an explicit gradient such as Nelder-Mead simplex minimization \citep{nelder-mead1965}.

A reasonable question one may ask is why use the $L_2$-norm as opposed to $L_1$ or some other similarity score.  A possible answer is that if we were to assume Gaussian noise, then the $L_2$-norm would be equivalent to the maximum likelihood estimation.  Another answer is that $L_1$ is a continuously differentiable function and hence is easier to compute, which allows for sensitivity analysis to more easily be conducted.  Finally, LS can be done without writing down explicit assumptions on the noise within the model.  However, we cannot guarantee properties of consistency or converging to a known distribution without more explicit assumptions.

On the other hand, least squares is easy to implement in software such as base \texttt{R}'s \texttt{optim()} function, and as we see in Section \ref{sec:results}, produces comparable methods to other methods.

@@

\begin{align*}
  \left \{
  \begin{array}{cl}
    \frac{dX}{dt} &= - \rr \gamma Y \frac{X}{N}\vspace{.5em}\\
    \frac{dY}{dt} &=  \rr \gamma Y \frac{X}{N}  - \gamma Y\vspace{.5em} \\
    \frac{dZ}{dt} &=  - \gamma Y 
  \end{array}
  \right . .
  \end{align*}
We find
\begin{align*}
(\hat{\rr}, \hat{\gamma} ) &= \text{argmin}_{\rr, \gamma} \sum_{t} \left [ \left (X_{obs}(t) - X(t)\right )^2 + \left ( Z_{obs}(t) - Z(t) \right )^2 \right ]
\end{align*}
We use the $\hat{\rr}$ directly from the above estimation process, which again can be done with a grid search or another optimization process.  It may be surprising that this method leads to different results than in simply using least squares.  One reason we obtain different results may be attributed to better numerical precision as there is one fewer step to obtain an estimate for $\rr$.  However, we also see that this method results in smaller standard errors than in LS, which we discuss more in Section \ref{sec:results}.

Like LS, while we have no explicit parametric assumptions on the noise, we also are uncertain of any theoretical guarantees of our resulting estimator.  Likewise, the same difficulties in sensitivity analysis arise as in LS.

@@

The SIR ODEs in Eq. \ref{eq:sir} have no known closed form solution, and so we are already using approximations using numerical integration.  In addition to this, data collected from real diseases are typically very noisy to begin with.  \cite{chang2017} discovered that an SIR model may be well approximated by a linear model.  We use this approach here to estimate $\rr$.

Specifically, we estimate two linear polynomials in \(t\) with degree \(K= 10\) to \(X_{obs}\)
and \(Y_{obs}\) using least squares to find the coefficients $\{(\hat{x}_k,
\hat{y}_k)\}_{k=1, \dots, K}$,
\begin{align*}
\hat{X}(t) &= \sum_{k=0}^K \hat{x}_k t^k\\
{\hat{Y}}(t) &= \sum_{k=0}^K \hat{y}_k t^k
\end{align*}
Then, we estimate the derivatives as
\begin{align*}
\hat{X}^\prime(t) &= \sum_{k=1}^K k \hat{x}_k t^{k-1}\\
\hat{Y}^\prime(t) &= \sum_{k=0}^K k \hat{y}_k t^{k-1}
\end{align*}
Following,  an estimator for \(\rr\) is derived from the ODEs in Equation \eqref{eq:sir},
\begin{align}
  - \frac{X^\prime}{Z^\prime}&= \rr \frac{X}{N} \nonumber\\
  \rr &=       -\frac{X^\prime}{
        Z^\prime} \cdot \frac{N}{X} \nonumber\\
  \hat{\rr} &= -\frac{\hat{X}^\prime(0)}{ \hat{Z}^\prime(0)} \cdot \frac{N}{\hat{X}(0)}. \nonumber
  \end{align}
  Here, $K=10$ is arbitrary and should be selected using some criterion such as AIC in addition with the knowledge of how many data points are available to the user.  Besides optionally deciding on the degree of polynomials to fit, this model is simple to implement and gives very comparable results to using least squares with the SIR model.  The time $t=0$ is used to best capture the initial outbreak.

  An advantage of using this estimation method is that it is simple to implement using any linear modelling software.  On the other hand, we are assuming $X(t) = \hat{X}(t)$ and so if the data truly follows a SIR model, then we will have biased estimators for $\rr$.  Using this method, we are able to check how sensitive $\rr$ is to the degree of the fitted polynomial.

@@

The above formulation (Section \ref{linear-model-approximation-degree-10}) of a linear model approximation only uses the estimate at time $t=0$ to estimate $\rr$.  We can instead, use all time points available to estimate $\rr$.  We fit a linear polynomial of \(t\) with degree \(K= 10\) model to \(X\)
and \(Z\) as above, with a slight modification in how we estimate
\(\rr\),
\begin{align*}
  \hat{\rr} &= \frac{1}{\# \text{ Obs }t} \sum_t \frac{-\hat{X}^\prime(t)}{\hat{Z}^\prime(t)} \cdot \frac{N}{X(0)} 
\end{align*}
The intuition is that $\frac{-X^\prime(t)}{Z^\prime(t)}$ is constant in $t$, but due to our approximations with the linear model, this is no longer the case.  Here, we average over the different possible values of $\rr$, estimated at different times.  An advantage to this approach is that we have a more robust estimate of $\rr$ than just using one time point.  However, the estimate will still be biased.  Like in LMAT, we can examine how sensitive $\rr$ is to the order of the polynomial $K$.

@@

Here we assume that the time step is small enough to approximate the incidence.  The advantage of this method is that incidence data is generally available as is prevalence data for certain diseases such as HIV \citep{hall2008}.  However, as one is required to have prior knowledge about $\gamma$, it may be easier to directly estimate $\rr$ with one of the many other methods described that does not require a prior knowledge about $\gamma$.  Again, we are using only one time point to estimate $\rr$.  This model is agnostic to assumptions on the noise, but again, it is unclear if this is an unbiased or consistent estimator of $\rr$.

\subsection{Smoothed Incidence to Prevalence Ratio (SIPR)}
We use the same method as above, IPR, but first estimate splines with 4 degrees of freedom, $\hat{X}(t)$ and $\hat{Y}(t)$, , to fit to $X_{obs}$ and $Y_{obs}$, respectively.  Then  $J(t) \approx -(\hat{X}(t+1) - \hat{X}(t))$, and the $\hat{\textnormal{IPR}}(t) = \frac{J(t)}{\hat{Y}(t)}$.  Then
\begin{align*}
\rr &= \hat{\textnormal{IPR}}(t) \cdot D
\end{align*}
The advantage of this method is that it creates a less noisy estimate than estimating IPR using only one point.  It has the same disadvantages as the regular IPR ratio in that it requires knowledge about $\gamma$.    Also, ratio estimators usually behave badly when the denominator is small.

@@

The advantages of this model are its simplicity of interpretation and ability to generate a whole probability distrubtion for $(X(t), Y(t))$.  We have explicitly stated how randomness enters the model so it is up to the researcher to check whether this assumption is a good fit.  However, it is possible to simulate how sensitive $\hat{\rr}$ is to $\hat{\alpha}$.

@@

>>>>

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
%\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}% http://ctan.org/pkg/graphicx
\usepackage{systeme}
%% tables
\usepackage{booktabs}
\usepackage{lscape}
\usepackage[table,xcdraw]{xcolor}
\usepackage{float}

@@

\section{Introduction}\label{sec:intro}
What has been called ``arguably the most important quantity in the study of epidemics'' \citep{Heesterbeek2002},  $\mathcal{R}_0$, the reproduction number (by convention pronounced ``R-naught'') is a persistently elusive quantity for epidemic modelers to estimate.  As defined by \citet{anderson1992}, $\rr$ is the ``the average number of secondary infections produced when one infected individual is introduced into a host population where everyone is susceptible.''  $\rr$, in some ways, summarizes an entire outbreak of a disease; it  is used to assess whether a disease epidemic will occur.  Additionally, it describes what percentage of the population needs to be vaccinated to avoid such an epidemic, roughly $1-\rr^{-1}$ \citep{anderson1992}).  Despite a clear definition of $\rr$, epidemiologists have struggled to create a standard  estimator for $\rr$  \citep{hethcote2000}.  A major issue in estimating $\rr$ is that the quantity is a \textit{property of the model}, meaning that $\rr$ is dependent not only on the usual noise that comes with statistical modeling but also on a variety of assumptions on how researchers assume a disease is transmitted through a population \citep{diekmann2009}.

To elaborate on $\rr$ being a property of the model, we first need to introduce the concept of the ``SI-framework,'' which was pioneered by Kermack and McKendrick in the 1920s \citep{getz2006}.   Here, S and I are known as compartments where S stands for ``susceptible'' and I for ``infectious.'' Infectious disease models then specify how individuals move from the S compartment to the I compartment.  Often, ancillary compartments are added such as R, which stands for recovered or dead, or E, which stands for exposed but not yet infectious.  The estimate of $\rr$ from a SIR model will not be directly comparable to the estimate of $\rr$ from a SEIR model, for example.  The difficulties of estimating $\rr$ are summarized and expanded upon by \cite{li2011}.   As a result of $\rr$ being a property of the model, we limit our review to methods for estimating $\rr$ for the SIR model introduced by \cite{Kermack700}.

@@

We review \wxxsir methods for estimating $\rr$.  We note that these methods are not exhaustive but are chosen due to their impact and use in epidemiology.  We find that these methods also help to highlight the problems discussed above.  In addition to making a point estimate of $\rr$, we are particularly interested in the hypothesis  $H_0: \rr > 1$, as a value greater than 1 indicates an outbreak of a disease.  As such, we discuss techniques for estimating CIs for each of the different methods.  These CI estimates are created using the delta method, the block bootstrap, and the posterior distribution.

In the following later, we present a series of simulations in which we examine each of our $\rr$ estimation methods.  Within these simulations, we vary the number of observations used to estimate $\rr$, infection and recovery rates, the total population size, initial SI ratio, and several assumptions about the noise.

Finally, we discuss our recommendations for estimating $\rr$ for the SIR model, and we comment on how these recommendations extend to more complex models and how many of the problems with estimating $\rr$  become exacerbated with the increase in complexity.

The rest of this manuscript is organized as follows.  In Section \ref{sec:r0}, we briefly discuss the origin of $\rr$.  In Section \ref{sec:sir-intro}, we introduce the SIR model as described by \cite{Kermack700}.  In Section \ref{sec:methods}, we overview the \wxxsir methods for estimating $\rr$. In Section \ref{sec:ci}, we discuss how to estimate CIs for each of the \wxxsir methods.  In Section \ref{sec:sim-res}, we present the results of our simulations.  Finally, in Section \ref{sec:discussion}, we provide recommendations of estimating $\rr$ for the SIR model and offer comments on how these conclusions extend to more complex models.

@@

Other difficulties in estimating $\rr$ arise from model assumptions on how randomness enters the model and sensitivity of the estimate.  For example, the SIR model is unique in that two of its compartments, S and R are monotonic.  The number of susceptibles is non-increasing and the number of susceptibles is non-decreasing.  Models should enforce this monotonicity of the two compartments.  As a consequence, this restricts assumptions on how randomness enters the model.  For example, Gaussian error, a common assumption in statistics and machine learning, becomes highly suspect due to the non-symmetrical nature of the noise imposed by the monotonic $S$ and $R$ compartments.  Without assuming Gaussian noise, it becomes much more difficult to guarantee properties such as unbiased and consistent estimates of $\rr$.

@@

Below in Table \ref{tab:r0-real-ex}, we provide examples of estimates $\rr$ from publications over the past three decades, spanning various locations over the world.  We display estimates for HIV, Zika (ZIK-V), Ebola (EVD), seasonal influenza, and H1N1 influenza.  These estimates of $\rr$ are not limited to those of SIR model, although most results come from this framework.  Of these, Zika seems to have the largest estimated values of $\rr$, followed by Ebola, H1N1 influenza, seasonal influenza, and HIV.  We see that African locations seem to have larger values of $\rr$ compared to the rest of the world, regardless of the disease.  Some of the accompanying lower and upper bound intervals (not necessarily CIs) are extremely large such as the case of HIV in Uganda (Interval: [.12, 14.17]) where as some are extremely small such as for Ebola in Guinea [1.5, 1.52].  We see that estimates for H1N1 influenza are generally higher than that of seasonal influenza, which is evidence that H1N1 was a bigger concern than the typical seasonal influenza cycle.  The lowest possible value for $\rr$ reported is .12 and the largest is 14.17, both for HIV in Uganda.  Within diseases, we see that estimates for $\rr$ seem to be similar, although typically do not lie in another's lower and upper bound, suggesting that there is great disparity in estimates of $\rr$.

% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
% \usepackage{graphicx}
% \usepackage{lscape}
\begin{landscape}
\centering
\begin{tabular}{@{}lllrrrl@{}}
\toprule
\textbf{Disease}    & \textbf{Location}                & \textbf{Year}                 & \textbf{$\rr$} & \textbf{Lower} & \textbf{Upper} & \textbf{Source}                                                                                                                  \\ \midrule
HIV                 & Belgium                          & 2004-2010                     & 1.10          & 1.09           & 1.16           & \cite{coelho2011}   \\
HIV                 & Netherlands                      & 2004-2010                     & 1.11        & 1.10            & 1.18           & \cite{coelho2011}    \\
HIV                 & Portugal                         & 2004-2010                     & 1.08        & 1.06           & 1.15           & \cite{coelho2011}   \\
HIV                 & San Francisco, USA               & 1993                          & 5.00           & 2.00              & 8.00              & \cite{blower1994}    \\
HIV                 & Uganda                           & 2002-2008                     & 3.31        & 0.12           & 14.17          & \cite{nsubuga2014}    \\ \hline
ZIK-V                & French Polynesia                 & 2013                          & 2.75        & 2.53           & 2.98           & \cite{zhang2017}        \\
ZIK-V                & French Polynesia                 & 2013-2014                     & 3.70         & 2.60            & 4.80            & \cite{kucharski2016}  \\
ZIK-V                & Florida                          & 2016                          & 0.16        & 0.13           & 0.19           & \cite{dinh2016}       \\
ZIK-V                & Colombia                         & 2015-2016                     & 2.50         & 3.00              & 3.60            & \cite{nishiura2016}    \\
ZIK-V                & Barranquilla, Colombia           & 2015                          & 2.40         & 3.80            & 5.60            & \cite{towers2016}    \\
ZIK-V                & Colombia                         & 2015                          & 1.42        & 2.56           & 3.83           & \cite{majumder2016}  \\ \hline
EVD               & Western Africa                   & 2014                          & 1.78        & 1.60            & 2.00              & \cite{fisman2014}   \\
EVD & Western Africa                   & 2014                          & 1.60         & 1.40            & 1.80            & \cite{towers2014}  \\
EVD                 & Democratic Republic of the Congo & 1995                          & 2.70         & 1.90            & 2.80            & \cite{legrand2007}  \\
EVD                 & Uganda                           & 2000                          & 2.70         & 2.50            & 4.10            & \cite{legrand2007}  \\
EVD                 & Guinea                           & 2014                          & 1.51        & 1.50            & 1.52           & \cite{althaus2014}   \\
EVD                 & Sierra Leone                     & 2014                          & 2.53        & 2.41           & 2.67           & \cite{althaus2014}      \\
EVD                 & Liberia                          & 2014                          & 1.59        & 1.57           & 1.60            & \cite{althaus2014}    \\ \hline
Seasonal Influenza  & USA, France, Australia           & 1972-2002 & 1.30         & 1.20            & 1.40            & \cite{chowell2008}  \\
Seasonal Influenza  & French Territory                 & 1987-1995                     & 1.50         & 1.09           & 1.73           & \cite{bonabeau1998}  \\ \hline
H1N1 Influenza      & Thailand                         & 2009                          & 2.07        & 1.92           & 2.22           & \cite{desilva2009}     \\
H1N1 Influenza     & Peru                             & 2009                          & 1.37        & 1.20            & 1.70            & \cite{desilva2009}        \\
H1N1 Influenza      & La Gloria, Mexico                  & 2009                          & 1.58        & 1.34           & 1.60            & \cite{fraser2009}     \\ \bottomrule
\end{tabular}
\caption{Estimates of $\rr$ from various researchers for the infectious diseases of HIV, ZIK-V, Ebola (EVD), seasonal influenza, and H1N1 pandemic influenza ranging from the years 1972-2016 in various locations.  We also report the lower and upper bounds reported by the researchers.}
\label{tab:r0-real-ex}
\end{table}
\end{landscape}

@@

In this particular  SIR model, $N$, the number of individuals is constant, as we see no arrows pointing away from the three nodes in Figure \ref{fig::sir}. Simple adaptations of the SIR model can include birth and death rates, which may correspondingly change the derivation of $\rr$ (for further discussion, see Section \ref{sec:discussion}). The parameters  $\beta$ and $\gamma$ have epidemiological meaning.  Here, $\beta$ is the average infection rate and $\gamma$ is the average recovery rate.  The movement of individuals from one compartment to another is represented through the ordinary differential equations below.  Both $\beta$ and $\gamma$ are constrained to be between $0$ and $1$.  For the remainder of this paper, we use $X$, $Y$, and $Z$ to denote the number of individuals in the $S$, $I$, and $R$ compartments, respectively, to avoid any confusion with $\rr$.
\begin{align}
\systeme{\frac{dX}{dt} = -\frac{\beta XY}{N}, \frac{dY}{dt} = \frac{\beta XY}{N} - \gamma Y, \frac{dZ}{dt} = \gamma Y}. \label{eq:sir}
\end{align}
In words, susceptible individuals become infected at a rate that is proportional to the percentage of infected individuals multiplied by $\beta$, the infection rate, and the number of susceptible individuals.  Infectious individuals recover at a rate of $\gamma$ multiplied by the number of infected individuals.

@@

In summary, when we discuss estimators for $\rr$ for the SIR model, we mean to say we are forming an estimator of $\rr$ from the given set of data in Eq. \ref{eq:sir-noise},
\begin{align*}
  \textnormal{Data} &= \left \{\left (X_{obs}(t), Y_{obs}(t), Z_{obs}(t) \right ) : t=0, 1, \dots, T\right \}, \\
  \hat{\rr} &= m(\textnormal{Data}),
\end{align*}
where $m$ is a function of the data.

Good properties of an estimator may include (1) being a consistent estimator
\begin{align*}
  \hat{\rr} \overset{P}{\to} \rr \textnormal{ as } T\to \infty
\end{align*}
or (2) converging to a known probability distribution,
\begin{align*}
\hat{\rr} \overset{d}{\to} F.  
\end{align*}
As we are interested in the hypothesis
\begin{align*}
  H_0:\;& \rr > 1 \\
  H_A:\;& \rr \le 1
\end{align*}
Then for some $\alpha \in [0,1]$, we would like to know for some $a, b >0$
\begin{align*}
P(a \le \hat{\rr} \le b) = 1 - \alpha,  
\end{align*}
the second property is particularly desirable, as we could find $a$ and $b$ from the distribution $F$.  At any rate, we would like to be able to estimate $a$ and $b$ for a given $\alpha$-level.  We could also make similar hypothesis tests such as that the $\rr$ for H1N1 influenza is greater than the reported $\rr$ for seasonal influenza.

@@

\subsection{Least Squares ($\beta$, $\gamma$) (LS)}\label{least-squares-beta-gamma}
Our first approach to estimate $\rr$ in the SIR model is to minimize the joint mean square error for the data collected at each time point and use the plug-in estimator found in Equation \ref{eq:sirls}.  In particular, we find

\begin{align*}
(\hat{\beta}, \hat{\gamma} )&=\arg \min_{\beta, \gamma} \sum_{t} \left [ \left (X_{obs}(t) - X(t)\right )^2 + \left ( Z_{obs}(t) - Z(t) \right )^2 \right ]
\end{align*}
Then the least squares (LS) estimate for $\rr$ is given by Equation \ref{eq:sirls},
\begin{align}\label{eq:sirls}
  \hat{\rr}= \frac{\hat{\beta}}{\hat{\gamma}}.
\end{align}

This method is sometimes used to estimate $\hat{\beta}$ and $\hat{\gamma}$ via a grid search of $\beta$ and $\gamma$ but can also be done with more sophisticated optimiization algorithms.  Since we cannot explicitly write down the partial derivatives of $S$, $I$, and $R$ with respect to $\beta$ and $\gamma$, we can only use minimization algorithms which do not rely on an explicit gradient such as Nelder-Mead simplex minimization \citep{nelder-mead1965}.

A reasonable question one may ask is why use the $L_2$-norm as opposed to $L_1$ or some other similarity score.  A possible answer is that if we were to assume Gaussian noise, then the $L_2$-norm would be equivalent to the maximum likelihood estimation.  Another answer is that $L_2$ is a continuously differentiable function and hence is easier to compute, which allows for sensitivity analysis to  be conducted more easily.  Finally, LS can be done without writing down explicit assumptions on the noise within the model.  However, we cannot guarantee properties of consistency or convergence to a known distribution without more explicit assumptions.

On the other hand, least squares is easy to implement in software such as \texttt{R}'s \texttt{optim()} function, and as we see in Section \ref{sec:results} produces comparable methods to other methods.  Least squares and similar variants have been used to estimate $\rr$ in \cite{majumder2016}.

@@

\begin{align*}
  \left \{
  \begin{array}{cl}
    \frac{dX}{dt} &= - \rr \gamma Y \frac{X}{N}\vspace{.5em}\\
    \frac{dY}{dt} &=  \rr \gamma Y \frac{X}{N}  - \gamma Y\vspace{.5em} \\
    \frac{dZ}{dt} &=  - \gamma Y 
  \end{array}
  \right . .
  \end{align*}
We find
\begin{align*}
(\hat{\rr}, \hat{\gamma} ) &= \text{argmin}_{\rr, \gamma} \sum_{t} \left [ \left (X_{obs}(t) - X(t)\right )^2 + \left ( Z_{obs}(t) - Z(t) \right )^2 \right ]
\end{align*}
We use the $\hat{\rr}$ directly from the above estimation problem, which again can be solved with a grid search or another optimization process.  It may be surprising that this method leads to different results than simply using least squares.  One reason we obtain different results may be attributed to better numerical precision as there is one fewer step to obtain an estimate for $\rr$.  However, we also see that this method further results in smaller standard errors than in LS, which we discuss more in Section \ref{sec:results}.

Like LS, while we have no explicit parametric assumptions on the noise, we also are uncertain of any theoretical properties of our resulting estimator.  Likewise, the same difficulties in sensitivity analysis arise as in LS.

@@

The SIR ODEs in Eq. \ref{eq:sir} have no known closed form solution, and so we use numerical integration to obtain approximate solutions.  In addition to this, data collected from real diseases are typically very noisy to begin with.  \cite{chang2017} showed that an SIR model may be well approximated by a linear model.  We use this approach here to estimate $\rr$.

Specifically, we estimate two linear polynomials in \(t\) with degree $K$  to \(X_{obs}\)
and \(Z_{obs}\) using least squares to find the coefficients $\{(\hat{x}_k,
\hat{z}_k)\}_{k=1, \dots, K}$,
\begin{align*}
\hat{X}(t) &= \sum_{k=0}^K \hat{x}_k t^k\\
{\hat{Z}}(t) &= \sum_{k=0}^K \hat{z}_k t^k
\end{align*}
Then, we estimate the derivatives as
\begin{align*}
\hat{X}^\prime(t) &= \sum_{k=1}^K k \hat{x}_k t^{k-1}\\
\hat{Z}^\prime(t) &= \sum_{k=0}^K k \hat{z}_k t^{k-1}
\end{align*}
Following,  an estimator for \(\rr\) is derived from the ODEs in Equation \eqref{eq:sir},
\begin{align}
  - \frac{X^\prime}{Z^\prime}&= \rr \frac{X}{N} \nonumber\\
  \rr &=       -\frac{X^\prime}{
        Z^\prime} \cdot \frac{N}{X} \nonumber\\
  \hat{\rr} &= -\frac{\hat{X}^\prime(0)}{ \hat{Z}^\prime(0)} \cdot \frac{N}{\hat{X}(0)}. \nonumber
  \end{align}
  Here, $K$ is arbitrary and should be selected using some criterion such as AIC.  Besides optionally deciding on the degree of polynomials to fit, this model is simple to implement and gives comparable results to using least squares with the SIR model.  The time $t=0$ is used to best capture the initial outbreak.

  An advantage of using this estimation method is that it is simple to implement using any linear modelling software.  On the other hand, we are assuming $X(t) = \hat{X}(t)$ and so if the data truly follows a SIR model, then we will likely have a biased result for $\rr$.  Using this method, we are able to check how sensitive $\rr$ is to the degree of the fitted polynomial.

@@

The above formulation (Section \ref{linear-model-approximation-degree-10}) of a linear model approximation only uses the estimate at time $t=0$ to estimate $\rr$.  We can instead, use all time points $O$ available to estimate $\rr$.  We fit a polynomial in \(t\) with degree \(K\) to \(X_{obs}\)
and \(Z_{obs}\) as above, with a slight modification in how we estimate
\(\rr\),
\begin{align*}
  \hat{\rr} &= \frac{1}{O} \sum_t \frac{-\hat{X}^\prime(t)}{\hat{Z}^\prime(t)} \cdot \frac{N}{X(0)} 
\end{align*}
The intuition is that $\frac{-X^\prime(t)}{Z^\prime(t)}$ is constant in $t$, but due to our approximations with the linear model, this is no longer the case.  Here, we average over the different possible values of $\rr$, estimated at different times.  An advantage to this approach is that we have a more robust estimate of $\rr$ than just using one time point.  Like in LMA, we can examine how sensitive $\rr$ is to the order of the polynomial $K$.

@@

Here we assume that the time step is small enough to approximate the incidence.  The advantage of this method is that incidence data is generally readily available as is prevalence data for certain diseases such as HIV.  However, as one is required to have prior knowledge about $\gamma$, it may be easier to directly estimate $\rr$ with one of the many other methods described that does not require a prior knowledge about $\gamma$.  Again, we are using only one time point to estimate $\rr$.  This model is agnostic to assumptions on the noise, but again, it is unclear if this is an unbiased or consistent estimator of $\rr$.

\subsection{Smoothed Incidence to Prevalence Ratio (SIPR)}
We use the same method as above, IPR, but first estimate splines with $K$ degrees of freedom, where $K$ is small, $\hat{X}(t)$ and $\hat{Y}(t)$, to fit to $X_{obs}$ and $Y_{obs}$, respectively.  Then  $J(t) \approx -(\hat{X}(t+1) - \hat{X}(t))$, and the $\hat{\textnormal{IPR}}(t) = \frac{J(t)}{\hat{Y}(t)}$.  Then
\begin{align*}
\rr &= \hat{\textnormal{IPR}}(t) \cdot D
\end{align*}
The advantage of this method is that it creates a less variable estimate than estimating IPR using only one point.  It has the same disadvantages as the regular IPR ratio in that it requires knowledge about $\gamma$.

@@

The advantages of this model are its simplicity of interpretation and ability to generate a whole probability distrubtion for $(X(t), Y(t))$.  We have explicitly stated how randomness enters the model so it is up to the researcher to check whether this assumption is a good fit.  Also, it is possible to simulate how sensitive $\hat{\rr}$ is to $\hat{\alpha}$.

@@

	Total population $N$: 100000.
